Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-06-10 paper/manuscript.tex

% aux. comments Stefan Boresch -> Stefan Boresch
\newcommand{\sbnote}[1]{%
  {\bfseries{}[SB: }%
  {\textcolor{blue}{#1}}{\bfseries{}]}
}

% aux. comments Emilio Gallicchio
\newcommand{\egnote}[1]{%
  {\bfseries{}[EG: }%
  {\textcolor{green}{#1}}{\bfseries{}]}
}

\newcommand{\rref}{{\bfseries[REFs]}{}}

@@

\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed, meaning there is no MCSS match between the two compounds.}
    \label{fig:fig_mcss}
\end{figure}

There are different tools that allow the generation of MCSS matches as well as single topology input. A large number of software tools can compute MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}. The MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space. Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode. If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.

Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem. Using an absolute free energy approach instead seems more promising in such a case.

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.

@@

An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species. Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested. Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units. A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).

In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex; Eq.~(\ref{eq:intra-zRL-def}).  The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average. To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom. Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand. To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
\begin{equation} \label{eq:Vsite-def}
    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
\end{equation}
%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
%for dynamical variables rather than integrated quantities
which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent. The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart. The result is
\begin{equation}\label{eq:DG0-binding-constant-def-average}
K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
\end{equation}
where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
\begin{equation}\label{eq:pert-energy-def}
\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
\end{equation}
is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble. Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
\end{equation}
which, expressed in free energy units, yields
\begin{eqnarray}\label{eq:DG0-def-Vsite}
  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
\end{eqnarray}

The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$. The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration. The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$. As seen above, the ideal term is a fundamental aspect of the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.

Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system. The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.

\paragraph{Handling the standard state in free energy calculations.}
As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward. Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations. These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions. Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.

As discussed above, the situation for binding is more complex and requires special care. The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor. Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy. Moreover, the indicator function prescribes that the ligand remains in the binding site region throughout the alchemical process, including when it is uncoupled from the receptor. Consequently, alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates. Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below). Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.

Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions. The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs). However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$. Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}. Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions. Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined. The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.

\paragraph{Several choices of binding site restraints are possible.}
A variety of restraint types are commonly used in alchemical binding free energy calculations. Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations. Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process. In particular, they must be present unchanged at the unbound and bound states. In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies. Hence, they are typically turned on during an initial stage, starting from the uncoupled state, and are turned off in the final stage, near the bound state. In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing binding sites restraints. However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are instead auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.

Binding site restraints implement the $I(\zeta)$ indicator function. They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor. Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate. The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice. The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand. In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions. However, practical considerations may be important. For example, some choices of Boresch-style anchor atoms can be numerically unstable. Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail. Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.

Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
Fu et al. suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}. Salari et al. proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}. DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms. However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints. These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.

As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}. Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}). However, it might be challenging to fully explore the bound macrostate at all alchemical states when it is only limited by receptor-ligand positional restraints. As further discussed below, auxiliary restraints are particularly helpful in the latter circumstance to limit the extent of conformational sampling necessary to achieve rapid convergence of binding free energy estimates.

Auxiliary restraints take a variety of forms, such as simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use} as well as flat-bottom restraints. The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints. For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose. Auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values. However, more computational effort is required to compute the free energy when the restraints are turned on and off. Additionally, such restraints can limit the exploration of alternative binding modes. This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.

\subsection{Absolute and relative calculations deal with some of the same issues} 
\subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
\egnote{Is this section necessary? We could add a comment regarding weak binders in the preceding section clarifying how the theory covers weak binders (as well as non-binders) and the precautions the user should take to obtain accurate estimates (relative to the experiments).}
In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}). Determining what the bound states actually are can be challenging for weakly bound ligands. For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.

@@

In absolute calculations, the need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}),  while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}). \egnote{I do not agree with the latter. Binding site restraints apply to the coupled and uncoupled states. The same is true for RBFE for both ligands. (see above)}
In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.

>>>>

% aux. comments Stefan Boresch -> Stefan Boresch
\newcommand{\sbnote}[1]{%
  {\bfseries{}[SB: }%
  {\textcolor{blue}{#1}}{\bfseries{}]}
  }
\newcommand{\rref}{{\bfseries[REFs]}{}}

@@

\paragraph{Atom mapping}
Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping. This is the case especially for system representations using a single coordinate set (Type 1. & 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done. This however requires either careful implementation of atom mass changes, bond length etc. or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors. \cite{ries2024Kartograf}

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed, meaning there is no MCSS match between the two compounds.}
    \label{fig:fig_mcss}
\end{figure}

In general there are mutliple ways on how a MCSS can be calculated. One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
A large number of software tools can compute SIS MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry. Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}. The SIS MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.

Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules. These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g. stereochemistry or binding mode differing).
The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned. Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.

In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode. If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.

A next step for atom mappings, would be to support multistate methods (e.g. RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms. There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect. \cite{petrov2021Perturbation}

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation . This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}). Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, Rieder2022Lev}

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.

@@

Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.

Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
\begin{equation} \label{eq:DGfromKAB}
    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
\end{equation}
Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.

\paragraph{Handling the standard state in absolute free energy calculations.}
For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.

For binding, however, the situation is more complex and requires special care.
Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
\begin{equation}\label{eq:restraint-correction}
    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
\end{equation}
where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.

\paragraph{Several choices of restraints are possible.}\label{ABFErest}
In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.

Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.

In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
Additionally, such restraints would typically keep the ligand from exploring alternative binding modes. This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.

Many choices of restraints involve selecting reference atoms.
Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example. Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.

More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms. Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand. Fu et al. suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}. Salari et al. proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}. This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction. This necessitates an additional step where the DBC restraints are released to a single harmonic restraint. Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.

\subsection{Absolute and relative calculations deal with some of the same issues}
\subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}). Determining what the bound states actually are can be challenging for weakly bound ligands. For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.

@@

In absolute calculations, this need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.

However, this strategy of using kinetic trapping to maintain the bound complex can fail when the dissociation rate of the ligand has the same or smaller order of magnitude than the length of the simulation.
This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou2017pushing,pan2017quantitative}.
In the case of weak binders, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
We stress that this type of restraint is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
It is important to note that the spring constant and/or radius parameters of the restraint effectively determine which conformations are considered to be bound.
As a consequence, these parameters must be tuned to the system so that only the binding site is accessible to the ligand.
Again, this step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.

In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Sec.~\ref{sec:standardstate-restraints}).
In the latter case, however, care must be taken when computing the standard state correction.
When multiple restraining potentials are active in the non-interacting state, the correction can generally be computed only through numerical integration.
Alternatively, one can adopt a protocol that removes the bound-state restraint in the non-interacting state.
Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensional,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992new,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.
